Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2021-01-27 manu_draft.Rmd
back trans section added

@@

@@

For all results, model S was used with a sufficiently high number of knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South SFE.  This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)). Chlorophyll trends were generally well-explained across all stations (Table \@ref(tab:modprftab)), with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`). GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll (Figure \@ref(fig:prddat)). All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, followed by a decreasing trend until the end of the record. Seasonally, all models indicated a larger spring peak in chlorophyll, particularly for more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location (Figure \@ref(fig:prddat)).

>>>>

For all results, model S was used with enough knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South SFE.  This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).

@@

### Back-transformation of model results

Model results were back-transformed from log$_{10}$-space to aid in the interpretation of trends. Back-transformation was accomplished using a bias correction factor for model predictions, estimates of mean values, and endpoints of confidence intervals from GAM results.  Using a bias correction factor correctly back-transforms the values to arithmetic space, whereas a simple exponentiation of the results produces under-estimates at the scale of the geometric mean [@Bradu70;@Duan83]. The following equation was used for back-transformation, such that:

\begin{equation}
E[y] = 10^{\mu + \frac{\log(10) \sigma^2}{2}}
(\#eq:backtrans)
\end{equation}

where the back-transformed, expected value of the response variable $y$ (chlorophyll) is a function of the mean value $\mu$ in log$_{10}$-space, corrected by a dispersion estimate $\sigma^2$.  The dispersion is obtained directly from the fitted GAM object as an estimate of the variance of the residuals, corrected by the model degrees of freedom.

Slope estimates that were used to determine significance of a seasonal trend across years from the mixed-meta analysis were not back-transformed.  While it is possible to back-transform model predictions, estimates of the mean values, and endpoints of confidence intervals, it is not possible for slope estimates in log$_{10}$-space that can change in arithmetic space across the values of the independent variable.  However, an "approximate" linear slope was estimated for the mixed-meta regression models as the linear fit across the seasonal averages that were back-transformed using equation \@ref(eq:backtrans).  This was done solely for interpretation of relative change across years, whereas the significance of the slope parameter in log$_{10}$-space from the mixed-meta model was used to assess significance of the trend.

@@

Model predictions for chlorophyll trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`). GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chlorophyll towards the more southern stations (Figure \@ref(fig:prddat)). All the models suggested 1) increasing chlorophyll from 1990 until 2005 to 2010, followed by decreasing chlorophyll until the end of the record in 2019, 2) a spring chlorophyll peak, particularly at south Bay stations, and 3) a fall chlorophyll peak that was smaller than the spring peak.  The magnitude of the fall peak did not vary noticeably by location (Figure \@ref(fig:prddat)).
